### 2020-05-31 writeup/intro.tex
sec 1

\begin{abstract}
When does multi-task learning outperform single-task learning?
In this work, we address this question by studying the test performance of predicting a particular task given multiple tasks using a commonly used hard parameter sharing architecture.
We consider multi-task learning in the setting of high-dimensional linear regression.
First, we provide a sharp analysis of the bias-variance tradeoff for multi-task learning, and extend the result to the related setting of transfer learning.
We develop a technical tool on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$ for two random matrices $X_1, X_2$ in the setting of two tasks.
Second, we apply our tool to rigorously explain when multi-task learning outperforms single-task learning.
%And we theoretically analyze the benefit of multi-task learning for reducing the amount of labeled data needed to achieve comparable performance to single-task learning. %, which has been a key empirical finding in recent work.
Finally, we show practical implications of our theory for detecting and mitigating negative effects in image and text classification tasks.
\end{abstract}

@@

In this work, we consider the setting of high-dimensional linear regression and focus on predicting a particular task whose amount of labeled data is limited.
Following Hastie et al. \cite{HMRT19} and Bartlett et al. \cite{BLLT20}, we assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Let $n_i$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$, for every $1\le i\le t$.
The label of task $i$ is given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes the ground truth parameters for task $i$ and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
Without loss of generality, let the $t$-th task denote the target task.
We assume that the data size of task $t$ is a small constant $\rho_t > 1$ times $p$ to capture the need for more labeled data.
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.
We focus on the hard parameter sharing architecture with a linear shared layer $B\in\real^{p\times r}$ for all tasks and a separate prediction head $\set{W_i \in \real^{r}}_{i=1}^t$ for each task \cite{R17,MTDNN19,WZR20}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
For a target task $t$, let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal multi-task (MTL) estimator (defined precisely in Section \ref{sec_setup}).
Let $\hat{\beta}_t^{\STL}$ denote the standard single-task (STL) estimator for task $t$.
We say there is negative transfer if the prediction loss of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.

\textbf{Main results.}
We revisit the bias-variance tradeoff of the multi-task estimator.
Interestingly, we observe that the variance of $\hat{\beta}_t^{\MTL}$ is always smaller than that of $\hat{\beta}_t^{\STL}$, hence resulting in a positive effect of variance reduction.
The bias of $\hat{\beta}_t^{\MTL}$, which we term as \textit{model shift bias}, results in a negative effect caused by the difference between $\beta_t$ and the rest of $\set{\beta_i}_{i=1}^{t-1}$.
Hence, the tradeoff between the variance reduction effect and model shift bias determines whether the transfer is positive.

Our first contribution is to develop a new technical tool that provides a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, which arises naturally from the setting of two tasks.
Our result extends a well-known result on the trace of $(X_1^{\top}X_1)^{-1}$ for a single random matrix \cite{S07}.
Furthermore, we provide a nearly optimal error bound, which may be of independent interest.
We apply our tool to both multi-task and transfer learning settings.
We present the following results in Section \ref{sec_main}.
\squishlist
	\item We provide a sharp analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two settings:
(i) two tasks with general covariance matrice; (ii) any number of tasks that have the same features (e.g. images).
	\item We extend our result to transfer learning in the setting of high-dimensional linear regression.
	We pool learnt source task representations into a shared body similar to $B$ in the hard parameter sharing architecture, as in Taskonomy by Zamir et al.'18 \cite{ZSSGM18}.
	We show that the model shift bias can be captured by the projection of $\beta_t$ to the orthogonal subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
\squishend

>>>>

\begin{abstract}
When does multi-task learning outperform single-task learning?
In this work, we address this question by studying the performance of predicting a particular task given multiple tasks.
We consider a multi-task learning architecture that uses a shared layer for all tasks and a separate output layer for each task.
First, for the setting of high-dimensional linear regression, we provide a sharp analysis of the bias-variance tradeoff of multi-task learning estimators.
We develop a technical tool on the inverse of the sum of two random matrices that naturally arises from the analysis.
We extend the result to the related setting of transfer learning.
Second, we apply our tool to rigorously explain when multi-task learning outperforms single-task learning.
%And we theoretically analyze the benefit of multi-task learning for reducing the amount of labeled data needed to achieve comparable performance to single-task learning. %, which has been a key empirical finding in recent work.
Finally, we show practical implications of our theory for detecting and mitigating negative effects in image and text classification tasks.
\end{abstract}

@@

In this work, we consider the setting of high-dimensional linear regression and focus on predicting a particular task whose amount of labeled data is limited.
Following Hastie et al. \cite{HMRT19} and Bartlett et al. \cite{BLLT20}, we assume that for every task $1\le i\le t$, its features are random vectors $x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Let $n_i$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
Without loss of generality, let the $t$-th task denote the target task.
Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
Hence adding more labeled data can help improve its test performance.
%We shall assume that each task data follows a linear model, i.e. $y_i = X_i \beta_i + \varepsilon_i$, $1\le i\le k$.
%Here $\beta_i\in\real^p$ is the model parameter for the $i$-th task.
%Each row of $X_i\in\real^{n_i\times p}$ is assumed to be drawn i.i.d. from a fixed
%distribution with covariance matrix $\Sigma_i$.
We focus on the hard parameter sharing architecture with a linear shared layer $B\in\real^{p\times r}$ for all tasks and a separate prediction head $\set{W_i \in \real^{r}}_{i=1}^t$ for each task \cite{R17,MTDNN19,WZR20}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
Let $\hat{\beta}_t^{\MTL} = B W_t$ denote the optimal multi-task (MTL) estimator for the target task.
Let $\hat{\beta}_t^{\STL}$ denote the single-task (STL) estimator.
We define these precisely in Section \ref{sec_setup}.
We say there is negative transfer if the prediction loss of $\hat{\beta}_t^{\MTL}$  is larger than that of $\hat{\beta}_t^{\STL}$, or positive transfer otherwise.

\textbf{Main results.}
We revisit the bias-variance tradeoff of the multi-task estimator.
Interestingly, we observe that the variance of the multi-task estimator is always smaller than the variance of the single-task estimator, hence resulting in a positive effect of variance reduction.
The bias of the multi-task estimator results in a negative effect caused by the difference between $\beta_t$ and the rest of $\set{\beta_i}_{i=1}^{t-1}$.
Hence, the bias-variance tradeoff determines whether we observe positive or negative transfer.

Our first contribution is to develop a new technical tool for a quantity that arises naturally from the analysis of two tasks.
Concretely, we show a tight bound on the trace of $(X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}$, which extends a well-known result on the trace of $(X_1^{\top}X_1)^{-1}$ for the single-task setting \cite{S07}.
Our analysis yields nearly optimal concentration error, which may be of independent interest.
Using this tool, we show the following results in Section \ref{sec_main}.
\squishlist
	\item We provide a sharp analysis of the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two settings:
(i) two tasks with general covariance matrice; (ii) any number of tasks that have the same features (e.g. images).
	\item We extend our result to transfer learning in the setting of high-dimensional linear regression.
	We pool learnt source task representations into a shared body similar to $B$ in the hard parameter sharing architecture, as in Taskonomy by Zamir et al.'18 \cite{ZSSGM18}.
	We show that the model shift bias can be captured by the projection of $\beta_t$ to the orthogonal subspace spanned by $\set{\beta_i}_{i=1}^{t-1}$.
\squishend
